|

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group. Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\). The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment. If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment. A GAM version of a linear model can be written as

@@

In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}D. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth. It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline. The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.

With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A, where the four basis functions (and the intercept) are shown. Each of the five basis functions is evaluated at six different points (because there are six points on the timeline). The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model. The penalized coefficient estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.

To get the weighted basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values per group, which appear as points.

@@

\caption{Basis functions for a single smoother for time. \textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. \textbf{B}: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. \textbf{C}: Weighted basis functions. Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. \textbf{D}: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.}\label{fig:basis-plot}
\end{figure}

@@

Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}. Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples. Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}. Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.

In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}. With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.

@@

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

@@

When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}. A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.

@@

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.

@@

\caption{Pairwise comparisons for smooth terms. \textbf{A}: Pairwise comparisons for the full dataset. \textbf{B}: Pairwise comparisons for the dataset with missing observations. Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0. In both cases the effect of treatment is significant after day 3. In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp}
\end{figure}

The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment. This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).

Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B, D. Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference. Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared. Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons. This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A. However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic. However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.

In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other. Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\). This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model. Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups. However, because the model was still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.

In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}. For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant. This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.

@@

In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}. This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs. T

The purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs. In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.

However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented. We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.

One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate. The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
\CSLLeftMargin{32. }
\CSLRightInline{Rose NL, Yang H, Turner SD, Simpson GL. An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}. \emph{Geochimica et Cosmochimica Acta}. 2012;82:113-135. doi:\href{https://doi.org/10.1016/j.gca.2010.12.026}{10.1016/j.gca.2010.12.026}}

>>>>

where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group. Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by the smooth term \(f(time_{t}|\beta_{j})\). The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment. \footnote{If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment} A GAM version of a linear model can be written as

@@

In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 that appears in Figure \ref{fig:l-q-response}D. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth. It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline. The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. In other words, a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.

With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}. In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown. Each of the five basis functions is evaluated at six different points (because there are six points on the timeline). The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model. The penalized coefficient estimates fitted are shown in Figure \ref{fig:basis-plot}B.

To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions. Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values which appear as points.

@@

\caption{Basis functions for a single smoother for time. \textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. \textbf{B}: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. \textbf{C}: Weighed basis functions. Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. \textbf{D}: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.}\label{fig:basis-plot}
\end{figure}

@@

Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}. Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples. Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.

In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}. What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.

With this in mind, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends, and use simultaneous empirical Bayesian CIs to assess significance between treatment groups.

@@

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. A brief guide for model selection and diagnostics appears in Appendix A, and the code for the simulated data and figures can be found in Appendix B.

@@

When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}} as we show in Appendix A. Additional discussions on model selection can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.

@@

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.

@@

\caption{Pairwise comparisons for smooth terms. \textbf{A}: Pairwise comparisons for the full dataset. \textbf{B}: Pairwise comparisons for the dataset with incomplete observations. Significant differences exist where the 95\% empirical Bayesian simultaneous CI does not cover 0. In both cases the effect of treatment is significant after day 3. For the difference, we have included the means so the the value of the difference has direct correspondence with the scale of the response (Figure \ref{fig:sim-smooth-plot}A).}\label{fig:plot-pairwise-comp}
\end{figure}

The absence of a \emph{p-value} in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.

With this expectation of different trends in each group, computing the difference between the trends will identify if and \emph{when} the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group). Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere \emph{p-value} because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain \emph{the whole function} at a nominal value (say, 95\%).

Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D. Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference. Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared. Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons. This can be seen in Figure \ref{fig:sim-smooth-plot}B, where at day 5 there is essentially a difference of 50\% between \(\mbox{StO}_2\) in both groups, which corresponds to the -50\% difference in \ref{fig:plot-pairwise-comp}A. However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic. However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.

In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other. Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\). This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model. Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups. However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.

For biomedical studies, the ability of smooth comparisons to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.

@@

In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends. It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}. This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.

However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect. In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.

However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented. We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in Appendix A.

One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, thereby providing a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate. The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
\CSLLeftMargin{32. }
\CSLRightInline{Rose NL, Yang H, Turner SD, Simpson GL. An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to {S}cotland, {UK}. \emph{Geochimica et Cosmochimica Acta}. 2012;82:113-135. doi:\href{https://doi.org/10.1016/j.gca.2010.12.026}{10.1016/j.gca.2010.12.026}}

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.Rmd
Working on Appendix B
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@

@@

>>>>

thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))

@@

## Final considerations

In this Appendix we have shown a basic model selection workflow for GAMs. Our goal here is to familiarize the reader with the logic behind the construction of each model and the kind of diagnostic information that needs to be checked to ensure the model is adequate. One final consideration must be given to the conditional distribution of the response. In the main manuscript and through this workflow, we have assumed a normal distribution for the response, which we believe is appropriate in many biomedical scenarios. However, we must remind the reader that the normal distribution is not intended to be used as a "one size fits all" distribution. Depending on the type of data (counts, binary outcomes) the user can choose different conditional distributions in _mgcv_.

We also want to indicate that one of the major advantages of fitting GAMs in R using _mgcv_ is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using `k.check`. If that is  not the case, then the user needs to increase `k` a bit and check `k.check` again. The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.

@@

Finally, Noam Ross has assembled a list of online tools and resources for learning about and using GAMs in R, which cover a range of very useful videos, slides, and courses from different authors. The list can be found in Noam Ross's GitHub profile here:

[Resources for Learning About and Using GAMs in R](https://github.com/noamross/gam-resources)

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

#alpha for ribbon in the smooth plots
al<-0.8
\end{lstlisting}

@@

@@

>>>>

#alpha for ribbon in the smooth plots
al<-0.8

thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
\end{lstlisting}

@@

\hypertarget{final-considerations}{%
\subsection{Final considerations}\label{final-considerations}}

In this Appendix we have shown a basic model selection workflow for GAMs. Our goal here is to familiarize the reader with the logic behind the construction of each model and the kind of diagnostic information that needs to be checked to ensure the model is adequate. One final consideration must be given to the conditional distribution of the response. In the main manuscript and through this workflow, we have assumed a normal distribution for the response, which we believe is appropriate in many biomedical scenarios. However, we must remind the reader that the normal distribution is not intended to be used as a ``one size fits all'' distribution. Depending on the type of data (counts, binary outcomes) the user can choose different conditional distributions in \emph{mgcv}.

We also want to indicate that one of the major advantages of fitting GAMs in R using \emph{mgcv} is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using \passthrough{\lstinline!k.check!}. If that is not the case, then the user needs to increase \passthrough{\lstinline!k!} a bit and check \passthrough{\lstinline!k.check!} again. The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.

@@

Finally, Noam Ross has assembled a list of online tools and resources for learning about and using GAMs in R, which cover a range of very useful videos, slides, and courses from different authors. The list can be found in Noam Ross's GitHub profile here:

\href{https://github.com/noamross/gam-resources}{Resources for Learning About and Using GAMs in R}

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

The next chunk presents the code used to create Figure 2 in the main manuscript. Here, we simulate data from the same concave down quadratic trend from Figure \@ref(fig:quadratic-cases-Appendix)A and use the package `mgcv` to fit a GAM to fit the trend of the data over time. The GAM in this case can be represented as

 $$Response= \beta_0 + f(time_t|\beta_1)$$
 Because we only use the data from one group and we do not account for interaction. The model in `mgcv` is written as:

@@

Because we want to include the group means for plotting purposes, we extract the intercept (`const`) and add it to the Treatment group estimates in both `ci` and `si` order to shift the interval so it matches the scale of the response. Finally, we create a `ggplot2` object for plotting.

@@

In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM. The package `mgcv` has the function `gam.check` to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a `ggplot2` object. Therefore we create the function `gam_diagnostics` that uses the same code from `gam.check` but without the graphical output. In this way, we can use `appraise` from the package `gratia` to create a graphical output in ggplot2 format and the quantitative information from `gam.check`.

>>>>

The next chunk presents the code used to create Figure 2 in the main manuscript. Here, we simulate data from the same concave down quadratic trend from Figure \@ref(fig:quadratic-cases-Appendix)A and use the package _mgcv_ to fit a GAM to fit the trend of the data over time. The GAM in this case can be represented as

 $$Response= \beta_0 + f(time_t|\beta_1)$$
 Because we only use the data from one group and we do not account for interaction. The model in _mgcv_ is written as:

@@

Because we want to include the group means for plotting purposes, we extract the intercept (`const`) and add it to the Treatment group estimates in both `ci` and `si` order to shift the interval so it matches the scale of the response. Finally, we create a _ggplot2_ object for plotting.

@@

In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM. The package _mgcv_ has the function `gam.check` to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a _ggplot2_ object. Therefore we create the function `gam_diagnostics` that uses the same code from `gam.check` but without the graphical output. In this way, we can use `appraise` from the package _gratia_ to create a graphical output in _ggplot2_ format and the quantitative information from `gam.check`.
